Imputation of Incomplete Multilevel Data with R

### 2023-12-11 docs/manuscript.qmd
Add complete case study 1 and 2, and text of 3

---
title: "Imputation of Incomplete Multilevel Data with R"
title-formatted: "Imputation of Incomplete Multilevel Data with [R]{.proglang}"
format:
    jss-pdf:
        keep-tex: true
        fig-pos: 't'
        fig-width: 7
        fig-height: 3.5
    jss-html: default
author:
  - name: Hanne I. Oberman
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CH
      - Journal of Statistical Software
    orcid: 0000-0003-3276-2141
    email: h.i.oberman@uu.nl
    url: https://www.hanneoberman.github.io
  - name: Johanna Muñoz
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-2384-5415
  - name: Valentijn M.T. de Jong
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        city: Utrecht
        country: The Netherlands
      - name: European Medicines Agency
        department: Data Analytics and Methods Task Force
        city: Amsterdam
        country: The Netherlands
      - Journal of Statistical Software
    orcid: 0000-0001-9921-3468
  - name: Gerko Vink
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0001-9767-1924
  - name: Thomas P.A. Debray
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-1790-2719    
abstract: |
  This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} packackage [mice]{.pkg}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed. This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.
keywords: [missing data, multilevel, clustering, mice, R]
keywords-formatted: [missing data, multilevel, clustering, "[mice]{.pkg}", "[R]{.proglang}"]
bibliography: bibliography.bib  
---

@@

```{r}
#| label: setup
#| echo: false
library(ggplot2)
source("../docs/utils.R")
```

@@

```{R}
#| echo: false
concepts <-
  data.frame(
    Concept = c(
      "Sample units",
      "Cluster",
      "Hierarchical data",
      "Level-1",
      "Level-2",
      "Hierarchical model",
      "Fixed effect",
      "Random effect",
      "Mixed effect",
      "ICC",
      "Stratified intercept"
    ),
    Details = c(
      "Units of the population from which measurements are taken in a sample, e.g., students.",
      "Variable that specify the cluster or agruppation, e.g., Classroom",
      "Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.",
      "Variable that varies within a cluster, eg. Test score",
      "Variable that does not vary within a cluster but between, e.g. teacher experience.",
      "Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)" ,
      "Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)",
      "Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. test score tendency during academic year between students due to no controlled factors such as  genetic,family history",
      "Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).",
      "The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
""
    )
  )
# concepts |>
#   kable(
#     booktabs = T,
#     escape = F,
#     caption = "Concepts in multilevel methods",
#     col.names = c("Concept", "Details"),
#     align = "l"
#   ) |>
#   column_spec(c(1), width = "3cm") |>
#   column_spec(c(2), width = "12cm")
```

@@

```{R}
#| echo: false
plot_na()
```

@@

::: callout
Box 3. The predictor matrix. The entries corresponding to the level-1 predictors are coded with a 3, indicating that both the original values as well as the cluster means of the predictor are included into the imputation model. The entry of 4 in the predictor matrix adds three variables to the imputation model for the imputation model predictor: the value of the predictor, the cluster means of the predictor and the random slopes of the predictor.
:::

@@

```{R}
#| prompt: true
#| warning: false
#| message: false
set.seed(123)
library(mice)
library(ggmice)
library(ggplot2)
library(miceadds)
library(lme4)
library(mitml)
library(broom.mixed)
```

@@

```{R}
#| prompt: true
data("popmis", package = "mice")
```

```{R}
#| prompt: true
dat <- popmis[, c("school", "teachpop", "popular", "texp", "sex")] 
```

```{R}
#| fig-cap: "Polar axis plot"
#| fig-alt: "A line plot on a polar axis"
ggmice(dat, aes(popular, teachpop)) + 
  geom_jitter()
```

With the `ggmice` unction `plot_pattern` we can visualize this.

```{R}
#| prompt: true
#| label: fig-pattern
#| fig-cap: "Missing data pattern."
#| fig-pos: 't'
plot_pattern(dat)
```

```{R, eval=FALSE}
#| layout-ncol: 2
#| echo: false
# plot_pattern(dat)
# plot_corr(dat)
```

```{R}
#| prompt: true
#| fig-cap: "Pair-wise correlations."
plot_corr(dat)
```

@@

@@

Evaluate the distribution of imputed values.
```{R}
ggmice(imp, aes(popular, group = .imp)) + 
  geom_density() 
```

Evaluate the distribution of imputed values.
```{R}
ggmice(imp, aes(.imp, popular)) + 
  geom_jitter(alpha = 0.05) +
    geom_boxplot()
```

```{R}
ggmice(imp, aes(popular, teachpop)) + 
  geom_jitter() +
  facet_wrap(~ .imp)
```

Analyze the imputed data.
```{R}
fit <- with(
  imp,
  lmer(teachpop ~ popular + texp + (1 | school))
)
```

Pool the estimates.
```{R}
pool(fit)
```

Display results in table.
```{R}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

>>>>

---
title: "Imputation of Incomplete Multilevel Data with R"
title-formatted: "Imputation of Incomplete Multilevel Data with [R]{.proglang}"
format:
    jss-pdf:
        keep-tex: true
        fig-pos: 'h'
        fig-width: 7
        fig-height: 2.5
    jss-html: default
author:
  - name: Hanne I. Oberman
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CH
      - Journal of Statistical Software
    orcid: 0000-0003-3276-2141
    email: h.i.oberman@uu.nl
    url: https://www.hanneoberman.github.io
  - name: Johanna Muñoz
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-2384-5415
  - name: Valentijn M.T. de Jong
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        city: Utrecht
        country: The Netherlands
      - name: European Medicines Agency
        department: Data Analytics and Methods Task Force
        city: Amsterdam
        country: The Netherlands
      - Journal of Statistical Software
    orcid: 0000-0001-9921-3468
  - name: Gerko Vink
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0001-9767-1924
  - name: Thomas P.A. Debray
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-1790-2719    
abstract: |
  This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} packackage [mice]{.pkg}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed. This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.
keywords: [missing data, multilevel, clustering, mice, R]
keywords-formatted: [missing data, multilevel, clustering, "[mice]{.pkg}", "[R]{.proglang}"]
bibliography: bibliography.bib  
---

@@

In hierarchical datasets, clustering is a concern because the homoscedasticity in the error terms cannot be assumed across clusters and the relationship among variables may vary at different hierarchical levels. When multiple imputation is used to deal with missing data, as the imputation and analysis process is performed separately, it is necessary that imputation model being congenial with the main analysis model (Meng, 1994), e.g. if the main model accounts for the hierarchical structure also imputation model should do it (Audigier, 2021). Not including clustering into the imputation process may lead to effect estimates with smaller standard errors and inflated type I error.

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
library(knitr)
library(ggplot2)
source("../docs/utils.R")
source("../docs/tables.R")
```

@@

```{R}
#| echo: false
# concepts |>
#   kable(
#     booktabs = T,
#     escape = F,
#     caption = "Concepts in multilevel methods",
#     col.names = c("Concept", "Details"),
#     align = "l"
#   ) |>
#   column_spec(c(1), width = "3cm") |>
#   column_spec(c(2), width = "12cm")
```

@@

```{R}
#| echo: false
#| fig-height: 1.5
plot_na()
```

@@

::: callout
Box 3. The predictor matrix. The entries corresponding to the level-1 predictors are coded with a 3, indicating that both the original values as well as the cluster means of the predictor are included into the imputation model. The entry of 4 in the predictor matrix adds three variables to the imputation model for the imputation model predictor: the value of the predictor, the cluster means of the predictor and the random slopes of the predictor.
- -2 = cluster variable 
- 1 = overall effect
- 3 = overall + group-level effect 
- 4 = individual-level (random) and group-level (fixed) effect 
:::

## Multilevel imputation workflow {#sec-workflow}

There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.

The selection of each strategy depends mainly on the assumptions in the main analysis and also on the restriction of the analyzed data.

Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009). The last restriction is also required when imputations are performed on each cluster separately.  When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).

Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster. However, this strategy is best suited for balanced data (Grund, 2017) and when random effects model is appropriated, i.e. the number of clusters is adequate. (Austin,2018).

Here it is important to evaluate the assumptions imposed by the main model, for instance by using the cluster indicator strategy may lead to bias estimates when the model is based on a hierarchical model (Taaljard,2008). Even when an imputation strategy congenial with the main model is preferred, it is important to consider whether it is appropriate for the data as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios(Bailey 2020). For instance, in causal effect analysis, separately imputation may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions. (Zhang,2023).

Below we provide a imputation workflow that can be used in general to impute cluster data.

### Focus on the Main Analysis
When imputing clustered data, we initially focus on the research question and the intended analysis, assuming the presence of non-incomplete values. This assessment sheds light on research hypotheses and the contemplated main(s) statistical model(s) and it provides insights into the data's structure (refer to the level table), variable types (e.g., confounders, auxiliary variables), and other considerations about variables relationships such as interactions or polynomial terms.

### Exploration of Available Data
Next, it is necessary to explore the information available in the data. Exploring individual variables using tools such as histograms or QQ plots can help to delineate variable distributions and plausible ranges of values and also identify input errors or outliers. 
Evaluating interactions between predictors using scatter plots or conditional plots, assessing collinearity using VIF or correlations, can help to glimpse nonlinear relationships that may affect the original formulation of the main model. Further testing of assumptions related to the response variable, including variance (homogeneity using conditional box plots), independence (e.g., ACF, variograms) and response vs. predictor relationships (conditional plots - x vs. y), may serve to choose an imputation model more suitable to the data.  In addition, the intraclass correlation coefficient (ICC) can be examined to assess cluster differences, aiding in the choice between the 2l and 1l methods for imputation.

One can also explore missingness, specifically missing patterns at the cluster level to identify types of patterns (e.g., non-monotonic, systematic) to guide the selection of an appropriate imputation approach in terms of computational efficiency (e.g., simpler regression imputation versus FCS in univariate patterns). Similarly, these missing patterns can also be used to identify potential predictors in individual imputation models by means of inflow criteria (quantifying the connections between missing data in one variable and other observed variables) and outflow criteria (determining the connections between observed values in one variable and missing data in other variables). 
Although several packages exist for visualizing missing data, this tutorial focuses on the recently implemented ggmice package.

### Assess Estimation Procedure Robustness to Missing Data
Prior to implementing multiple imputation, it is critical to assess the robustness of the estimation model when confronted with missing data. There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods. For instance, when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{Molenberghs_2007}. Furthermore, depending on the proportion of missingness (usually is <5%), opting for a simpler complete case analysis might be sufficient.

### Pre-imputation
Before proceeding with the imputation model, it is necessary to filter the dataset to include only the minimal variables required for the main model. If additional procedures are conducted during the analysis, it is essential to include variables associated with these procedures, such as confounders in the case of balancing techniques. Other relevant variables, like instrumental variables or auxiliary variables, may enhance parameter estimates even if not included in the main model, as they could be linked to the probability of missingness for some incomplete variables.

Furthermore, at the variable level, it is crucial to assess whether proxy variables should be considered or if deterministic imputation alone is adequate for imputing missing values. For instance, certain incomplete variables may not necessitate stochastic imputation methods like MICE. Instead, they can be effectively addressed through deductive imputation, where incomplete values are inferred from logical and deterministic relationships between variables. This approach is especially beneficial for variables that are functions of others, such as deriving a person's BMI from their weight and height. It also proves valuable for determining values for one-level variables from two-level ones, as seen in the context of Individual Participant Data (IPD). In such cases, incomplete information can be deduced from metadata, like inferring incomplete data about abortion in a country where abortion is illegal, or through cross-temporal or protocol-based deduction, such as imputing missing test values for deceased patients.

### Setting Imputation Model

##### Clustering Inclusion
Various strategies can be employed in the imputation process to account for clustering, including the inclusion of a cluster indicator variable, conducting a separate imputation process for each cluster, or employing a simultaneous imputation method that considers clustering\cite{eddings}

The choice of strategy depends primarily on the assumptions made in the main analysis and the constraints imposed by the analyzed data. Concerning data restrictions when the analysis do not contemplate an hierarchical model (eg. descriptive way), if there are few clusters with many observations per cluster, the cluster indicator or imputation separated by groups may be appropriate \cite{graham2009}. Conversely, an imputation model that simultaneously imputes all clusters using a hierarchical model can be employed when there are more clusters or fewer observations per cluster \cite{Allison_2002}.

In a hierarchical imputation model, correlations between observations within clusters are modeled by a random effect, allowing estimation even with a limited number of observations per cluster. Multiple imputation models based on hierarchical models have been proposed, each with different assumptions. \cite{audigier}

In the particular case that the analysis contemplates the use of a hierarchical model (i.e. main model), before choosing a particular strategy, it is crucial to verify if the assumptions of the imputation model align with those of the main analysis. For example, using the cluster indicator strategy may introduce bias estimates when the model is based on a hierarchical structure \cite{taljaard2008,speidel2018}. Even if an imputation strategy congruent with the main model is preferred, it is essential to assess its appropriateness for the data, considering that less complex imputation strategies may lead to unbiased estimates in certain scenarios \cite{bailey2020}.

#### Choice of Individual Imputation Methods
The initial step involves selecting the individual imputation model for each incomplete variable in the dataset. The mice package automatically proposes imputation methods based on the variable type for 1l variables. For 2l level variables, the micemd package's () function can be utilized. This function selects among the 2l imputation methods based on the size of the clusters and the proportion of missingness in each cluster.

In addition to the package-defined imputation methods, users can specify custom methods using the "I formula". This flexibility allows for the calculation of derived variables internally or adjustments to imputation methods based on specific conditions, such as conditioning the imputation model on the level of an incomplete covariate (e.g., pregnancy test on females).

#### Model Specification
The imputation model must be congenial with the main model \cite{meng1994}. Congeniality issues arise when the imputation model and the main model make different assumptions, often due to the omission of a polynomial or interaction term or the use of transformed variables.

The imputation model can include additional terms than the main model terms that do not lead to uncongeniality. For instance, it is advisable to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}. In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be included as a covariate in the imputation model [REF]. Furthermore, the inclusion of auxiliary variables, even if not part of the main model, can be associated with the probability of missingness, enhancing the likelihood of satisfying the Missing at Random (MAR) assumption and improving estimation efficiency \cite{hardt2012a}.

The specification of imputation models is done on a variable basis, either using a prediction matrix where the type of prediction variable should be specified accordingly (see table) or through a list of formulas in the formula parameter. The latter option proves useful in formulating imputation models with polynomial terms or interactions compared to the prediction matrix specification, which requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}.

However, even when considering an interaction term, depending on the adapted individual imputation model, it has been suggested, for instance, for treatment interaction effects, to conduct separated imputation by treatment group \cite{zhang2023}. Additionally, there are imputation models based on random forest or deep learning that can handle interaction and non-linear terms that do not require the explicit inclusion of the non-linear terms.

### Post-Imputation

During the imputation process, certain issues may arise that halt the process. In hierarchical model imputations, many issues are linked to methods based on parametric hierarchical models, which may struggle to estimate a model midway through the process. In such cases, it is advisable to examine the imputation log file to identify variables with imputation issues. Subsequently, reducing the number of predictors in the imputation model could be explored, either by using functions like quickpred or by considering variable transformations, such as scaling. Adjusting the level of the hierarchical model (e.g., using a homogenous variance model or a 1l model) can also be beneficial. Additionally, checking the range of imputed variables is crucial, as excessively large imputed values for one predictor could lead to convergence issues in other variables. This can be mitigated by including post-processing specifications on problematic variables or by using imputation models like Predicted Mean Matching (PMM), ensuring imputed values align with the observable values. In certain cases, a separate imputation strategy may be considered. For instance, in analyses involving multiple endpoints, conducting separate imputation processes for each endpoint might be preferable to a unified imputation process.

### Convergence and Sensitivity Analysis

Before starting into the analysis of each imputed dataset, it is crucial to validate the convergence of the imputation process. This is commonly accomplished through trace plots that depict the mean and variance of the incomplete variables across iterations. These plots serve to uncover potential circular issues or the need for additional iterations. Additionally,  it is also important to verify that imputed values fall within a plausible range and also to check the distribution of imputed variables, ensuring that the imputed variable distribution aligns with the distribution of observed values (under the MAR assumption). An alternative approach involves assessing the prediction accuracy of the imputation method \cite{cai2023}.

While the majority of Multiple Imputation by Chained Equations (MICE) methods are based on Missing at Random (MAR) assumptions, field expert input may suggest that the Missing Not at Random (MNAR) mechanism could be plausible for certain variables. An MNAR variable is one in which the probability of missingness depends on an unobservable variable. This can occur when missingness is associated with the incomplete value itself (self-marking) or when there is an unobserved variable linked to both the value and the probability of missingness of the incomplete value (indirectly non-informative). Specifically, for the indirectly non-informative case in hierarchical datasets, imputation methods based on the Heckman method can be considered.\cite{hammon2020,hammon2022,munoz2023}

---

@@

```{R}
#| prompt: true
#| warning: false
#| message: false
set.seed(123)         # for reproducibility
library(lme4)         # for multilevel modeling
library(mice)         # for imputation
library(miceadds)     # for multilevel imputation methods
library(micemd)       # for selection-model imputation methods
library(mitml)        # for multilevel parameter pooling
library(broom)        # for clean model estimates
library(broom.mixed)  # for multilevel model estimates
library(dplyr)        # for data wrangling
library(ggmice)       # for visualization
library(ggplot2)      # for visualization
```

@@

In this section we will go over the different steps involved with imputing incomplete multilevel data with the R package mice. We consider the simulated `popmis` dataset, which included pupils ($n = 2000$) clustered within schools ($N = 100$). The following variables are of primary interest:

  - `school`,     school identification number (clustering variable);
  - `popular`,    pupil popularity (self-rating between 0 and 10; unit-level);
  - `sex`,	      pupil sex (0 = boy, 1 = girl; unit-level);
  - `texp`,	      teacher experience (in years; cluster-level).

The analysis model corresponding to this dataset is multilevel regression with random intercepts for the different schools. We will estimate the association between the pupils' sex and their popularity score. This model can be expressed in `lme4` code as:

```r
popular ~ 1 + sex + (1 | school)
```

We load the data into the environment with
```{R}
#| prompt: true
data("popmis", package = "mice")
```

and select the relevant variables
```{R}
#| prompt: true
dat <- popmis[, c("school", "popular", "texp", "sex")] 
```

which results in the following data structure.
```{r}
#| prompt: true
head(dat)
```

The association of interest can be visualized with `ggmice`,
```{R}
#| fig-cap: "Scatterplot of student popularity by sex"
#| fig-alt: ""
ggmice(dat, aes(sex, popular)) +
  geom_jitter()
```
where missing datapoints in the `popular` variable are represented by red points on the X-axis of the figure.

With the `ggmice` function `plot_pattern` we can visualize the missing data pattern 
```{R}
#| prompt: true
#| label: fig-pattern
#| fig-cap: "Missing data pattern."
plot_pattern(dat)
```

which shows us that the missingness is univariate and sporadic.

To develop the best imputation model for the incomplete variable `popular`, we need to know whether the observed values of `popular` are related to observed values of other variables. Plot the pair-wise complete correlations in the incomplete data
```{R}
#| prompt: true
#| fig-cap: "Pair-wise correlations."
plot_corr(dat)
```

This shows us that  not just the analysis-model variable `sex`, but also the cluster-level covariate teacher experience, `texp`, may be a useful as an imputation model predictor. Moreover, the missingness in `popular` may depend on the observed values of other variables. With `ggmice()` we can visualize the distribution of the teacher experience for cases where `popular` is observed and cases where `popular` is missing.
```{r}
#| prompt: true
ggmice(dat, aes(texp)) + 
  geom_density() +
  facet_wrap(~is.na(popular))
```

It appears that students with a missing value for `popular` are in clusters with a slightly higher `texp` value.
```{r}
t.test(dat$texp ~ is.na(dat$popular)) |> 
  tidy() |>
  kable()
```

Although there are no significant differences in the distribution of `texp` depending on the missingness indicator of `popular`, this variable can serve as auxiliary variable in the imputation of `popular`.

@@

The `pmm` method is better (more efficient) because it will still look for donors (maybe outside of cluster) based on predictive distance, even for very small clusters.

@@

Troubleshoot non-convergence of imputation model. Evaluate the distribution of imputed values.
```{R}
#| prompt: true
ggmice(imp, aes(popular, group = .imp)) + 
  geom_density() 
```

Evaluate the distribution of imputed values.
```{R}
#| prompt: true
ggmice(imp, aes(.imp, popular)) + 
  geom_jitter(alpha = 0.05) +
    geom_boxplot()
```

```{R}
#| prompt: true
ggmice(imp, aes(sex, popular)) +
  geom_jitter() +
  facet_wrap(~ .imp)
```

Analyze the imputed data.
```{R}
#| prompt: true
fit <- with(
  imp,
  lmer(popular ~ texp + sex + (1 | school))
)
```

Pooling the estimates does not provide estimates of the variance components.
```{R}
#| prompt: true
pool(fit)
```

Therefore, `mitml` is used.
```{R}
#| prompt: true
est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

Display results in table.
```{r}
#| prompt: true
est$estimates |> 
  round(3) |>
  kable()
```

```{r}
#| prompt: true
est$extra.pars |> 
  round(3) |>
  kable()
```

### IMPACT data

The second case study is the `impact` data from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ units across $N = 15$ clusters, @metamisc].

The `impact` data set contains traumatic brain injury data on $n = 11022$ patients clustered in $N = 15$ studies with the following 11 variables:

  - `name` Name of the study,
  - `type` Type of study (RCT: randomized controlled trial, OBS: observational cohort),
  - `age` Age of the patient,
  - `motor_score` Glasgow Coma Scale motor score,
  - `pupil` Pupillary reactivity,
  - `ct` Marshall Computerized Tomography classification, 
  - `hypox` Hypoxia (0=no, 1=yes),
  - `hypots` Hypotension (0=no, 1=yes),
  - `tsah` Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  - `edh` Epidural hematoma (0=no, 1=yes),
  - `mort` 6-month mortality (0=alive, 1=dead).

Check if there is systematic missingness in this dataset. For illustration purposes, we made Marshall Computerized Tomography classification (ct) systematically missing.

The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model: 
```r 
mort ~ type + age + motor_score + pupil + ct + (1 | name) 
```

Note that variables `hypots`, `hypox`, `tsah` and `edh` are not part of the analysis model, and may thus serve as auxiliary variables for imputation.

The `impact` data included in the \pkg{metamisc} package is a complete data set. The original data has already been imputed once (Steyerberg et al, 2008). For the purpose of this tutorial we have induced missingness (mimicking the missing data in the original data set before imputation). The resulting incomplete data can be accessed from [zenodo link to be created](https://zenodo.com).

Load the incomplete data into the R workspace:
```{r data, eval = FALSE}
#| prompt: true
dat <- read.table("link/to/the/data.txt") 
```

```{r data-actually, include=FALSE}
#| prompt: true
dat <- readRDS("../Data/impact_incomplete.RDS")
```

To explore the missingness, we should look at the missing data pattern. The ten most frequent missingness patterns are shown with
```{r}
#| prompt: true
#| fig-height: 4
plot_pattern(dat, rotate = TRUE, npat = 10L)  
```

This shows that we need to impute `ct` and `pupil`. To develop the best imputation model, we need to investigate the relations between the observed values of the incomplete variables and the observed values of other variables, and the relation between the missingness indicators of the incomplete variables and the observed values of the other variables. To see whether the missingness depends on the observed values of other variables, we can test this statistically or use visual inspection (e.g. a histogram faceted by the missingness indicator).

We should impute the variables `ct` and `pupil` and any auxiliary variables we might want to use to impute these incomplete analysis model variables. We can evaluate which variables may be useful auxiliaries by plotting the pairwise complete correlations
```{r}
#| prompt: true
#| fig-height: 4
plot_corr(dat, rotate = TRUE)
```

This shows us that `hypox` and `hypot` would not be useful auxiliary variables for imputing `ct`. Depending on the minimum required correlation, `tsah` could be useful, while `edh` has the strongest correlation with `ct` out of all the variables in the data and should definitely be included in the imputation model. For the imputation of `pupil`, none of the potential auxiliary variables has a very strong relation, but `hypots` could be used. We conclude that we can exclude `hypox` from the data, since this is neither an analysis model variable nor an auxiliary variable for imputation
```{r}
#| prompt: true
dat <- select(dat, !hypox)
```

Mutate data to get the right data types for imputation (e.g. integer for clustering variable).
```{r}
#| prompt: true
dat <- dat |>
  mutate(name = as.integer(name))
```

Create an initial methods vector for the incomplete variables
```{r}
#| prompt: true
meth <- make.method(dat)
meth
```

which should be adjusted to the appropriate `2l` methods.
```{r}
#| prompt: true
meth[c("pupil", "ct")] <- "2l.pmm"
meth[c("hypots", "tsah", "edh")] <- "2l.pmm"
```

Create an initial predictor matrix 
```{r}
#| prompt: true
pred <- quickpred(dat)
```

This predictor matrix is too large to display inline. A visualization of the adapted predictor matrix is presented in Figure XYZ.

We should make sure `name` is used as clustering variable
```{r}
#| prompt: true
pred[, "name"] <- -2
```

and the analysis-model outcome should be used as a predictor in all imputation models
```{r}
#| prompt: true
pred[, "mort"] <- 2
```

the resulting predictor matrix is visualized with
```{r}
#| prompt: true
#| fig-height: 4
plot_pred(pred, method = meth, rotate = TRUE)
```

Impute the incomplete data with
```{r}
#| prompt: true
# imp <- mice(
#   dat,
#   method = meth,
#   predictorMatrix = pred,
#   printFlag = FALSE
# )
```

Evaluate the convergence of the algorithm
```{r}
#| prompt: true
# plot_trace(imp)
```

Analyze the imputed data:
```{r}
#| prompt: true
# fit <- imp %>% 
#   with(glmer(
#     mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name),
#     family = "binomial"
#     )) 
```
The estimated effects after imputation are presented in Table XYZ.

### obesity data

In this example, we demonstrate a multilevel imputation of random intercept and random slope model with a continuous response. We utilize the obesity dataset included in the `micemd`\@ package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries. We simulate data for 5 clusters so that the true values are known. We use the following variables from the dataset:

-   **Cluster:** Region of the patients' healthcare provider (Cluster variable),
-   **Gender:** Subjects' Gender (0=male, 1=female),
-   **Age:** Subjects' age,
-   **Height:** Subjects' height in metres,
-   **Weight:** Subjects' weight in kilograms,
-   **BMI:** Subjects' body mass index,
-   **FamOb:** Family obesity history (yes or no),
-   **Time:** Response time in minutes (exclusion-restriction variable).

In this dataset, Age and FamOb are MAR, while the weight variable is affected by selection bias, attributed to an indirect MNAR mechanism. This MNAR mechanism typically arises when an unobserved or omitted variable influences both the value of the incomplete variable (in this case, Weight) and its likelihood of being missing (denoted as R).

In the primary analysis model, BMI serves as the dependent variable, with Age, Gender, and FamOb as predictors. Because of the clustered nature of the data, which is quantified with the Intraclass Correlation Coefficient (ICC) below, we include random intercepts, as well as a random slope for the Age variable.
The model is represented as:
\begin{equation}
\label{eqn:main}
BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}
\end{equation}

We start by loading the data:
```{r}
#data("data_heckman", package = "micemd")
#dat <- data_heckman
```

Now, let's begin by examining the missing patterns in the data by cluster:

```{r, dpi = 300,out.width="70%", fig.cap="Missing pattern", eval=TRUE}
# myplots <- lapply(1:5, function(i) {
#   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
#   ggplot2::ggtitle(paste0("Cluster", i))
#  })
# ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
```

We observe that the missing pattern is non-monotonic and quite similar across the clusters. However, regarding the weight variable, we notice that is systematically missing in cluster 3. In order to evaluate
if we require a imputation method that accounts for clustering  we assess the Intraclass Correlation

```{r obesity-icc, eval=TRUE}
# Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
# performance::icc(Nulmodel)
```

Since the ICC is above 0.1 and as the main analysis will be use a mixed model, we decide
to use two-level (2l) imputation methods. In this imputation process, we include all predictor
variables from equation \ref{eqn:main} in the main model. However, since BMI is a composite of
weight and height, we use deterministic imputation for these, which is described below.

We use the **find.defaultMethodfunction** provided in the **micemd** package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters.

It suggests using '2l.2stage.bin' for the FAV variable and '2l.2stage.norm' for the age variable. However, after inspecting the age density plot, we consider modifying its method to
'2l.2stage.pmm'. For the BMI variable, we employ deterministic imputation.

```{r obesitymeth, eval=TRUE}
# meth_mar <- micemd::find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
#                                ni.small = 100, prop.small = 0.4)
# meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
# meth_mar["Age"]<-"2l.2stage.pmm" 
```
